---
title: "Practicals: session 13th"
subtitle: "Creating scenarios"
author: "Ali Abbas"
engine: knitr
editor: visual
toc: true
lightbox: true
format: 
  html: 
    theme: cosmos
    embed-resources: true
    toc: true
    code-tools: true
    code-fold: false
    page-layout: full
    df-print: paged
    number-sections: true
---

# Learning objectives:

In-depth look at the scenario generation with a better understanding of travel behaviour

```{r setup}
#| echo: false
#| warning: false
#| eval: true
#| message: false
library(tidyverse)
library(here)
library(gt)
library(plotly)
# Read io
io <- readRDS(here("results/multi_city/io.rds"))
# Read trips from the io object and remove ghost trips (with participant_id = 0)
trips <- io$bogota$trip_scen_sets |> filter(participant_id > 0)
# Set palette for gt table
global_palette <- "viridis" #c("#fee8c8", "#e34a33")


theme_set(theme_minimal(base_family = "Spline Sans"))
theme_update(
  panel.grid.minor = element_blank(),
  panel.grid.major = element_blank(),
  axis.line.x = element_line(color = "grey80", linewidth = .4),
  axis.ticks.x = element_line(color = "grey80", linewidth = .4),
  plot.margin = margin(10, 15, 10, 15)
)


scen_colours <- c("Baseline" = '#b15928',
                  "Cycling" = '#abdda4',
                  "Car" = '#d7191c',
                  "Bus" = '#2b83ba',
                  "Motorcycle" = '#fdae61')


mode_colours <- c("pedestrian" = '#99d8c9',
                  "cycle" = '#abdda4',
                  "car" = '#d7191c',
                  "bus" = '#2b83ba',
                  "motorcycle" = '#fdae61',
                  "other" = '#fee8c8')


```

# Baseline travel behaviour

## Trip distribution by mode

Before we get into scenario creation, let's have a close look at trips distribution in the travel behaviour

```{r}
#| warning: false
# Read in the trips dataset
trip_dist <- trips |>  
    # Keep unique trips id - ignoring stage level info
    distinct(trip_id, scenario, .keep_all = T) |> 
    # Remove ghost trips (vehicle trips)
    filter(participant_id > 0, 
           # Only look at the baseline travel diary
           scenario == "baseline") |> 
    # Count grouped data by trip mode
    count(trip_mode) |> 
    # Calculate frequency as proportion
    mutate(freq = n/sum(n)) |> 
    # Remove n
    dplyr::select(-n) |> 
    # Change to wide format
    pivot_wider(names_from = trip_mode, values_from = freq)

trip_dist |> 
    # Create a gt table
    gt() |> 
    # Format all numbers as percentage
    fmt_percent(
      columns =  where(is.numeric),
      decimals = 1
    ) |> 
  # Colour them according to their values
  data_color(columns = where(is.numeric), 
             # Pick range from min to max
             domain = c(trip_dist |> as.numeric() |> min(na.rm = T), 
                        trip_dist |> as.numeric() |> max(na.rm = T)), 
             # Set viridis palette
             palette = global_palette) |> 
  # Set table header and subtitle
  tab_header(
          title = "Percentage Distribution by Trip Mode",
          subtitle = "Values represent percentages summing to 100"
        )


```

This shows that `bus` share is the highest with `r round(trip_dist$bus * 100, 1)`(%) and `other` with the lowest share `r round(trip_dist$other * 100, 1)`(%).

## Trip distribution by distance categories

### Distribution using Cumulative Distribution Function (CDF)

We have defined three distance categories, which are:

1.  Short trips within `2 km` distance

2.  Medium trips between `2` and `6 km` distance

3.  Long trips with `6+ km` distance

They have been drawn from domain knowledge as well as from empirical distance distribution captured in many trip diary datasets. Do note that they are not fixed and can be user-defined in the parameters sheet. In this exercise, let's see how distances for `bus`, `car`, `cycle` and `pedestrian` trips are distributed in these distance categories.

```{r}
 plotly::ggplotly(ggplot(trips |> 
                           filter(scenario == "baseline", 
                                  stage_mode %in% c('bus', 'car', 
                                                    'cycle', 'pedestrian')), 
                         aes(x = stage_distance, color = stage_mode)) +
     stat_ecdf(geom = "step") +
     geom_vline(alpha = 0.3, xintercept = 2, linetype = "solid", color = "black") +
     geom_vline(alpha = 0.3, xintercept = 6, linetype = "solid", color = "black") +
     scale_colour_manual(values = mode_colours, name = "Travel Mode") +
     coord_cartesian(ylim = c(0, 1)) +
     labs(x = "Distance (km)", y = "Cumulative Distribution Function (CDF)",
          title = "Distribution of distance (km) by mode \n\n") +
     theme_minimal()) |> 
  plotly::add_annotations(text="2 km",
                  x = 1.2, 
                  y = 0, 
                  showarrow = TRUE,
                  textangle = -90, 
                  font=list(size = 10)) |> 
  plotly::add_annotations(text="6 km",
                  x = 5.2, 
                  y = 0, 
                  showarrow = TRUE,
                  textangle = -90, 
                  font = list(size = 10))
```

### Table with distance categories

```{r}
# Read in the trips dataset
trips_tbl <- trips |> 
  # Filter baseline trips
  filter(scenario == "baseline") |> 
  # Keep distinct trip - to ignore trips with multiple stages
  distinct(trip_id, .keep_all = T) |> 
  # Count trip_mode, scenario and trip_distance_cat
  count(trip_mode, scenario, trip_distance_cat) |>
  # Now group by trip_mode and scenario
  group_by(trip_mode, scenario) |>
  # Calculate proportion of each trip_distance_cat
  mutate(proportion = n / sum(n)) |>  
  # Remove count by trip_distance_cat
  dplyr::select(-n) |>
  ungroup() |>
  # Keep only complete rows
  complete(scenario, trip_mode, trip_distance_cat, fill = list(proportion = 0)) |> 
  # Convert from long to wide format
  pivot_wider(names_from = trip_distance_cat, values_from = proportion) 
  
trips_tbl |> 
  # Remove scenario as we only have baseline
  dplyr::select(-scenario) |> 
  # Create a gt table
  gt() |> 
  # Convert all numerics into percentages
  fmt_percent(
    columns =  where(is.numeric),
    decimals = 1
  ) |> 
  # Rename columns
  cols_label(
    trip_mode = "Travel Mode") |> 
  # Colour all numeric columns
  data_color(columns = where(is.numeric), 
             domain = c(0, 1), 
             palette = global_palette,
             apply_to = "fill") |> 
  # Add table headers
  tab_header(
    title = "Percentage Distribution by  Trip Distance Category",
    subtitle = "Values represent percentages summing to 100 per row"
  )
  
```

## How does it work under the hood?

### Example of `5%` increase in cycling mode

Assume that there are only two distance bands `A` and `B` and that `80%` of all `cycling` trips lie in distance band `A` and the remaining `20%` in distance band `B`. In terms of overall trips, assume that `60%` of all trips are in distance band `A` and `40%` in distance band `B`.

Let's assume we want to increase the `cycling` mode share by `5%` of all trips. Then, we need to convert:

-   `5% x 80% / 60% = 6.67%` of non-cycling trips in distance band `A` to cycling trips and

-   `5% x 20% / 40% = 2.5%` of non-cycling trips in distance band B to cycling trips.

-   Overall, this leads to an increase of `(5% x 80% / 60%) x 60%` + `(5% x 20% / 40%) * 40%` = `5%` of cycling trips, whilst preserving the cycling mode shares of `80%` in distance band `A` and `20%` in distance band `B` and preserving the total mode shares of `60%` in distance band `A` and `40%` in distance band `B`. The total number of trips is also preserved.

# Create a pedestrian scenario

In this exercise, let's create a scenario, independent of the `ithimr` package. Now that we have a good understanding of trip modes and their distances, let's modify some of the trips and calculate impact of it on physical activity. We will consider these:

-   the shortest distance category with distances less than `2` km

-   Create a new scenario with target mode as `pedestrian` trips

-   Do note that since `pedestrian` trips in the baseline account for the majority of trips already, as we saw in the previous exercise, so we shouldn't expect a substantial change in the PA volume.

## Create a copy of the `baseline` scenario

Using the `trips` dataset, let's create a copy of the `baseline` scenario. Fill in the scenario name below:

```{r}
# get baseline trips for 15-69 year old from the io object
trips_baseline <- trips |> 
  filter(scenario == "baseline", 
         participant_id > 0) 

# Assign them to a new variable called trips_ped_scen
trips_ped_scen <- trips_baseline
```

## Now change trips to `pedestrian` mode

Change all trips under `2 km` using the `trip_distance_cat` of `0-2km` to `pedestrian` mode.
Fill in the distance category and the mode of interest.

```{r}

# Filter all rows that have trip_distance_cat as `0-2km` and convert their mode to `pedestrian`
trips_ped_scen[trips_ped_scen$trip_distance_cat  == '0-2km' & 
                 trips_ped_scen$trip_mode != "pedestrian"]$trip_mode <- 'pedestrian'

# Do the same for `stage_mode` as well
trips_ped_scen[trips_ped_scen$trip_mode == "pedestrian"]$stage_mode <- 'pedestrian'

```

## Recalculate duration

Calculate duration using walking speed and distance (which doesn't change when trips are switched to `pedestrian` mode)

```{r}

WALKING_SPEED <- 2.5

# For all the `pedestrian` trips, calculate duration using
# duration in mins = distance in kms * 60 / WALKING_SPEED
trips_ped_scen[trips_ped_scen$trip_mode == "pedestrian"]$stage_duration <- trips_ped_scen[trips_ped_scen$trip_mode == "pedestrian"]$stage_distance * 60 / 2.5

```

## Assign a new name to the `pedestrian` scenario

Assign a name to the scenario. Use the same nomenclature as previous exercises with `sc_` followed by scenario name.

```{r}
trips_ped_scen$scenario <- 'sc_ped'
```

## Combine the two trip datasets

Create a new trips dataset where we join both `baseline` and `pedestrian` scenarios together. Use `rbind` function.

```{r}

ntrips <- rbind(trips_baseline, trips_ped_scen)

head(ntrips)
```

## Calculate duration of active travel for each participant

Calculate duration of active travel for each participant using the newly created `ntrips` dataset. Active travel modes are: `cycle` and `pedestrian`

```{r}
active_travel_duration <- ntrips |> 
  mutate(
      # Rename walk_to_pt to pedestrian
      stage_mode = case_when(grepl("walk_to_pt", stage_mode) ~ 
                                  "pedestrian", TRUE ~ stage_mode), 
      # Chage stage_duration in minutes to hours            
      stage_duration = stage_duration / 60) |> 
  # Group by participant_id for each scenario
  group_by(participant_id, scenario) |> 
  # Calculate overall duration per person per week spent in active travel
  reframe(dur_cyc_hrs = sum(stage_duration[stage_mode == "cycle"]),
          dur_ped_hrs = sum(stage_duration[stage_mode == "pedestrian"]))
# Show top rows
head(active_travel_duration)

```

## Convert duration to mMET hours per week

Using duration and marginal MET hours per week per walking and cycling, calculate travel related PA volume

```{r}
# Initialize walking and cycling MMET hours per week
WALKING_MMET <- 2.5
CYCLING_MMET <- 5.8

travel_mmets <- 
  # Use active travel duration per person dataset
  active_travel_duration |> 
  # Group by participant_id and scenario
  group_by(participant_id, scenario) |> 
  # calculate total travel MMET hours per week by multiplying duration of cycling and walking
  # by their mMET hours per week and summing them together.
  reframe(total_travel_mmets = dur_cyc_hrs * (CYCLING_MMET) + 
            dur_ped_hrs * (WALKING_MMET)) 

head(travel_mmets)

```

## Bring in background PA volume in mMET hours per week from `base_pop`

Bring background PA volume from `base_pop` dataset

```{r}
# Create mmet per person variable that brings in background PA volume from `base_pop` dataset
mmets_pp <- travel_mmets |> 
  ## join it with participant id and bring the `work_ltpa_marg_met` column
  ## that stores background PA for each individual
  left_join(io$bogota$base_pop |> dplyr::select(participant_id, work_ltpa_marg_met))

```

## Combine travel and background PA volume

Similar to the previous exercises, let's sum travel and background PA volumes together

```{r}

# Join travel and background PA volume together
mmets_pp_total <- mmets_pp |> 
  mutate(total_mmets = total_travel_mmets + work_ltpa_marg_met )
```

## Change the structure to `wide` format

Change the structure of the dataset from `long` to `wide` format. Use `names(mmets_pp_total)` on your console to find out which variable name stores the total mMET hours per week.

```{r}
# Read in the total mMET hours per week and change its orientation from long to wide format
mmets_pp_total_wide <- mmets_pp_total |> 
  # Remove columns that we are not interested in
  dplyr::select(-c(total_travel_mmets, work_ltpa_marg_met)) |> 
  # Take values from scenario and the variable that holds total mmets
  pivot_wider(names_from = scenario, values_from = total_mmets)

head(mmets_pp_total_wide)
```

## Plot distribution of the two scenarios

Plot the distribution of both scenario and see where the change has happened. 

```{r}
# Use ggplot create a CDF plot of both baseline and pedestrian scenarios
# Use plotly to make it an interactive plot 
plotly::ggplotly(
  mmets_pp_total |> 
    # Use total mmet variable as x and scenario as y 
    ggplot(aes(x = total_mmets, color = scenario)) + 
    stat_ecdf(geom = "step") +
    labs(title = "Cumulative Distribution of PA volume in mMET hours per week for baseline and pedestrian scenarios",
          x = "mMET hours per week", y = "CDF"))

```
