---
title: "Practical: Smoking-Attributable IHD Deaths in England, 2012–2023"
author: "Student name:"
date: "Date:"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

---

# Introduction

This practical has two questions:

> **Question 1.** What is the burden of smoking-attributable IHD mortality for males and
> females aged 50–59 in England, from 2012 to 2023?

> **Question 2.** How many IHD deaths could have been averted under a scenario of
> accelerated smoking decline?

You will use the **Population Impact Fraction (PIF)** formula for a binary exposure
(smoker vs non-smoker):

$$PIF = \frac{\sum_i p_i RR_i \;-\; \sum_i p_i^* RR_i}{\sum_i p_i RR_i}$$

With smoking as a binary exposure, in R this is one line:

```r
p      <- c(p_s, 1 - p_s)   # observed: smoker, non-smoker proportions
p_star <- c(0,   1)          # counterfactual: no one smokes
RR     <- c(RR_s, 1)         # relative risks (non-smoker = 1)

PIF <- (sum(p * RR) - sum(p_star * RR)) / sum(p * RR)
```

All data are provided below — no external files are needed.

---

# Data

## Smoking prevalence

Current cigarette smokers (%) in 50–59 year olds, England.  
Source: ONS Opinions and Lifestyle Survey, 2024.

```{r data-smoking}
smoking <- data.frame(
  year       = 2012:2023,
  prev_men   = c(21.9, 19.2, 18.4, 17.1, 15.9, 19.1, 14.7, 19.7, 16.3, 15.9, 17.7, 15.4),
  prev_women = c(17.4, 18.0, 17.3, 15.2, 13.2, 14.9, 15.9, 14.0, 15.6, 13.3, 12.2, 12.6)
)

smoking$p_men   <- smoking$prev_men   / 100
smoking$p_women <- smoking$prev_women / 100

smoking[, c("year", "prev_men", "prev_women")]
```

## IHD deaths

IHD deaths, England, ages 50–59 (50–54 + 55–59 summed), by sex.  
Source: GBD 2023 Results Tool (IHME).

```{r data-ihd}
ihd <- data.frame(
  year         = 2012:2023,
  deaths_men   = c(2884, 2914, 2855, 2915, 2948, 3033, 3131, 3173, 3477, 3568, 3632, 3882),
  deaths_women = c( 683,  713,  694,  724,  731,  740,  761,  774,  814,  853,  900,  959)
)

ihd
```

## Relative risks

Current smoker vs non-smoker, IHD, ages 45–59.  
Source: Ezzati et al., *Circulation* (2003), CPS-II reanalysis.

```{r data-rr}
RR_men   <- 3.04
RR_women <- 3.78
```

---

# Question 1 — Burden of smoking-attributable IHD mortality

## Step 1: Calculate the PIF for each year

Complete the blanks (`___`) in the code below. The loop runs once per year. For each
year, build the observed distributions `p_men_obs` and `p_women_obs`, then apply the
PIF formula in one line for each sex.

```{r q1-pif}
p_star <- c(0, 1)   # counterfactual: no one smokes

results <- data.frame(
  year       = 2012:2023,
  PIF_men    = NA_real_,
  PIF_women  = NA_real_,
  attr_men   = NA_real_,
  attr_women = NA_real_
)

for (i in 1:nrow(results)) {

  # Observed distributions: c(smoker proportion, non-smoker proportion)
  p_men_obs   <- c(smoking$p_men[i],   1 - smoking$p_men[i])
  p_women_obs <- c(smoking$p_women[i], 1 - smoking$p_women[i])

  # RR vectors: c(RR for smokers, RR for non-smokers)
  RR_men_vec   <- c(RR_men,   1)
  RR_women_vec <- c(RR_women, 1)

  # PIF — one line each
  results$PIF_men[i]   <- (sum(___ * ___) - sum(p_star * ___)) / sum(___ * ___)
  results$PIF_women[i] <- (sum(___ * ___) - sum(p_star * ___)) / sum(___ * ___)

  # Attributable deaths
  results$attr_men[i]   <- results$PIF_men[i]   * ihd$deaths_men[i]
  results$attr_women[i] <- results$PIF_women[i] * ihd$deaths_women[i]
}

results$PIF_men    <- round(results$PIF_men   * 100, 1)
results$PIF_women  <- round(results$PIF_women * 100, 1)
results$attr_men   <- round(results$attr_men)
results$attr_women <- round(results$attr_women)
results$attr_total <- results$attr_men + results$attr_women

results
```

## Step 2: Plot attributable deaths over time

```{r q1-plot, fig.width=8, fig.height=5}
plot(results$year, results$attr_men,
     type = "b", pch = 16, lwd = 2, col = "#1E3A5F",
     ylim = c(0, 1400),
     xlab = "Year",
     ylab = "Smoking-attributable IHD deaths",
     main = "Smoking-attributable IHD deaths, ages 50–59, England")

lines(results$year, results$attr_women,
      type = "b", pch = 17, lwd = 2, lty = 2, col = "#2A9D8F")

lines(results$year, results$attr_total,
      type = "b", pch = 15, lwd = 2, lty = 3, col = "#E76F51")

legend("topleft",
       legend = c("Men", "Women", "Total"),
       col    = c("#1E3A5F", "#2A9D8F", "#E76F51"),
       lwd = 2, lty = c(1, 2, 3), pch = c(16, 17, 15), bty = "n")
```

> **Q1.1** Smoking prevalence fell between 2012 and 2023 for both sexes. Yet total
> attributable IHD deaths increased. Explain why.

> **Q1.2** Women have a higher RR (3.78) than men (3.04), yet fewer attributable deaths
> in every year. What drives this difference?

> **Q1.3** Calculate the cumulative total of attributable IHD deaths across all 12 years.
> What does this figure represent, and what assumptions does it rest on?

```{r q1-cumulative}
cat("Cumulative attributable IHD deaths, 2012–2023:\n")
cat("  Men:   ", sum(results$attr_men),   "\n")
cat("  Women: ", sum(results$attr_women), "\n")
cat("  Total: ", sum(results$attr_total), "\n")
```

---

# Question 2 — Averted deaths under accelerated decline

## The scenario

Smoking prevalence in England has declined since 2012, but what if it had declined
*twice as fast*? The scenario sets the **counterfactual 2023 prevalence** to half the
**observed 2012 prevalence**:

| | Observed 2012 | Observed 2023 | **Scenario 2023** |
|---|---|---|---|
| Men | 21.9% | 15.4% | **10.95%** |
| Women | 17.4% | 12.6% | **8.70%** |

In the PIF formula: **$p$** remains the *observed 2023 distribution*; **$p^*$** becomes
the *scenario distribution*. The PIF then measures the fraction of deaths that would have
been averted by this faster decline.

## Step 3: Calculate averted deaths for 2023

```{r q2-scenario}
# Observed 2023 prevalences
p_men_2023   <- smoking$p_men[smoking$year   == 2023]
p_women_2023 <- smoking$p_women[smoking$year == 2023]

# Scenario counterfactual: half of 2012 prevalence
p_men_cf   <- smoking$p_men[smoking$year   == 2012] / 2
p_women_cf <- smoking$p_women[smoking$year == 2012] / 2

# 2023 IHD deaths
deaths_men_2023   <- ihd$deaths_men[ihd$year   == 2023]
deaths_women_2023 <- ihd$deaths_women[ihd$year == 2023]

# Observed distributions (2023 actual)
p_obs_men   <- c(p_men_2023,   1 - p_men_2023)
p_obs_women <- c(p_women_2023, 1 - p_women_2023)

# Scenario counterfactual distributions
p_cf_men   <- c(p_men_cf,   1 - p_men_cf)
p_cf_women <- c(p_women_cf, 1 - p_women_cf)

RR_men_vec   <- c(RR_men,   1)
RR_women_vec <- c(RR_women, 1)

# PIF comparing 2023 observed → scenario counterfactual
PIF_men_scenario   <- (sum(p_obs_men   * ___) - sum(___ * ___)) / sum(p_obs_men   * ___)
PIF_women_scenario <- (sum(p_obs_women * ___) - sum(___ * ___)) / sum(p_obs_women * ___)

attr_men_scenario   <- round(PIF_men_scenario   * deaths_men_2023)
attr_women_scenario <- round(PIF_women_scenario * deaths_women_2023)

# Compare to observed attributable deaths in 2023
attr_men_2023   <- results$attr_men[results$year   == 2023]
attr_women_2023 <- results$attr_women[results$year == 2023]

averted_men   <- attr_men_2023   - attr_men_scenario
averted_women <- attr_women_2023 - attr_women_scenario

cat("--- 2023 results ---\n")
cat("Attributable deaths (observed prevalence) — Men:", attr_men_2023,     " Women:", attr_women_2023,     "\n")
cat("Attributable deaths (scenario prevalence) — Men:", attr_men_scenario, " Women:", attr_women_scenario, "\n")
cat("Deaths averted                            — Men:", averted_men,       " Women:", averted_women,       "\n")
cat("Total averted:", averted_men + averted_women, "\n")
```

## Step 4: Plot observed vs scenario

```{r q2-plot, fig.width=7, fig.height=5}
heights <- c(attr_men_2023, attr_men_scenario, attr_women_2023, attr_women_scenario)
cols    <- c("#1E3A5F", "#7BA7C7", "#2A9D8F", "#8FD4CA")
labels  <- c("Men\n(observed)", "Men\n(scenario)", "Women\n(observed)", "Women\n(scenario)")

bp <- barplot(heights,
              names.arg = labels,
              col = cols, border = NA,
              ylab = "Smoking-attributable IHD deaths",
              main = "Observed vs. accelerated-decline scenario, 2023\nEngland, ages 50–59",
              ylim = c(0, max(heights) * 1.25))

text(x = bp, y = heights + 15, labels = heights, cex = 0.9)
```

> **Q2.1** How many IHD deaths would have been averted in 2023 if smoking had declined
> twice as fast since 2012? Express this as a percentage of total IHD deaths in this age
> group in 2023.

> **Q2.2** In the scenario, $p^*$ is set to the counterfactual prevalence. Those who
> "stop smoking" under the scenario are treated as equivalent to non-smokers (RR = 1).
> Is this realistic? In which direction does it bias the averted-deaths estimate?

> **Q2.3** Name one other factor that changed between 2012 and 2023 that this analysis
> does not account for.

---

# References

1. Office for National Statistics. *Adult Smoking Habits in England, 2024.* Newport: ONS; 2025.
2. GBD 2023 Causes of Death Collaborators. *Global Burden of Disease Study 2023.* Seattle: IHME; 2024.
3. Ezzati M, Lopez AD, Rodgers A, Vander Hoorn S, Murray CJL. Selected major risk factors and global and regional burden of disease. *Lancet.* 2002;360(9343):1347–1360.
